---
title: "LivingNorwayR Workshop"
author: "Matt"
date: "10 12 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r}
# load libraries -----------------------------------------------
library(tidyverse)
library(rgdal)
library(here)
library(devtools)
library(leaflet)
library(htmlwidgets)
library(htmltools)
library(webshot)

```



# Download a data set from GBIF

```{r}
datasetID <- "4a00502d-6342-4294-aad1-9727e5c24041"
dataset <- RJSONIO::fromJSON(paste0("http://api.gbif.org/v1/dataset/",datasetID,"/endpoint"))
# The data set is available as a Darwin COre Archive zip file
# from the "endpoint";
endpoint_url <- dataset[[1]]$url
dir.create("data")
ifelse(file.exists("data/TOVE.zip"), NA,
download.file(endpoint_url, destfile="data/TOVE.zip", mode="wb"))
```

```{r}
unzip("data/TOVE.zip",exdir="data") 
```

```{r}
#devtools::install_github("livingNorway/livingNorwayR")
library(LivingNorwayR)
```

```{r LNImport}
# Create a DwCArchive object from the downloaded Darwin Core archive file
TOVEArchive <- initializeDwCArchive("data/TOVE.zip", "UTF-8")
```

```{r}
TOVEArchive
```

```{r RetriveCoreTable}
# Retrieve the core table from the archive object
TOVEEventTable <- TOVEArchive$getCoreTable()
# TOVEEvent table is an object of type GBIFEvent which is derived from DwCGeneric
class(TOVEEventTable)
# Export the contents of the event table to a data frame
TOVEEventTableDF <- TOVEEventTable$exportAsDataFrame()
# Lets look at the top few rows of the data frame extracted from the core data table
head(TOVEEventTableDF)
```
```{r}
TOVEOccTable <- TOVEArchive$getExtensionTables(1)[[1]]
TOVEOccTable

```


```{r}
library(tidyverse)
TOVEOccTableDF <- TOVEOccTable$exportAsDataFrame()

CinclusTOVE=TOVEOccTableDF %>% 
  dplyr::filter(genus=="Cinclus")
```

```{r}
CinclusCount=CinclusTOVE %>% 
  select(organismQuantity, eventID, )

```

```{r}
MergeOccEvent=TOVEEventTableDF %>% 
  left_join(CinclusCount)

```
```{r}
library(sp)
MergeOccEvent=MergeOccEvent %>% 
  select(decimalLatitude, decimalLongitude, organismQuantity,stateProvince, municipality)

spatialCinclus=MergeOccEvent
coordinates(spatialCinclus)<-~decimalLongitude+decimalLatitude
plot(spatialCinclus)
```


```{r}
shape <- readOGR(dsn = paste0(here(),"/SpatialData"), layer = "gadm36_NOR_1")

shape@data$VARNAME_1

```



# map each region to a management region ----------------------------------

shape@data$MAN_REGION<-c("R4", "R4", "R2","R2", "R8", "R5", "R1", 
                         "R6", "R6", "R7", "R3", "R4", "R1", "R1",
                         "R6", "R2", "R8", "R1", "R2")

# plotting a leaflet map --------------------------------------------------

# get the colour pallette -------------------------------------------------
n = length(unique(shape@data$MAN_REGION))
factorPal <- colorFactor(viridis::viridis(n), shape@data$MAN_REGION)


# plot leaflet map --------------------------------------------------------

leaflet() %>% 
  addTiles() %>% 
  addPolygons(data=shape, weight = 2, color = "black",  
              fillColor = ~factorPal(shape@data$MAN_REGION),
              fillOpacity = 1,
              #label = ~shape@data$MAN_REGION)
              label = ~shape@data$MAN_REGION,
              labelOptions = labelOptions(noHide = F)) 

require(tidyverse)
require(viridis)
require(ggthemes)
require(plyr)

map.df <- fortify(shape, region ="MAN_REGION")



head(map.df)


ggplot() +  
  geom_polygon(data=map.df, aes(x=long, y=lat, group=group,fill=id), 
            color="grey50", size=0.25) +
  scale_fill_colorblind() +
  coord_equal() +
  theme_map() +
  theme(legend.position="bottom") +
  theme(legend.key.width=unit(2, "cm"))+
  labs(fill = "Management Regions")



